---
title: "Calculate Exposure Inferences:  Examples"
author:  "Zachary Stanfield"
date:  "February 11, 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calculateExpInferences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This vignette demonstrates how to utilize the full bayesmarker package pipeline from obtaining the NHANES data (https://wwwn.cdc.gov/nchs/nhanes) to inferring parent chemical exposures.  To limit required compuation time and data requirements,
the first part of this vignette analyzes a single NHANES cohort (2015-2016) and a single population group.  
Multiple cohorts can be run individually via loop (see section "Looping over multiple cohorts") or together by combining 
sampling data (see section "Combining Cohorts").


# Quick start:  single cohort and single population
Obtain chemical exposure estimates for the 2015-2016 NHANES cohort for children aged 0-5 years.

## Prepare for session
### Load the relevant librarires
```{r setup1}
library(bayesmarker)
```

### Assign needed function inputs 
The codes_file has specific formatting (see the example file included in this package or the manuscript supplemental data). 
```{r inputs1}
cohort <- "15-16"
codes_file <- "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx"
save_directory <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
data_path <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
```

## Use the bayesmarker package 
### Obtain and save the raw data
This call will result in a folder called "2015-2016" at the location "save_directory/rawData/".
The file will include all need files corresponding to the table in the codes file related to 2015-2016 metabolites.
```{r 1516children}
get_NHANES_data(codes_file = codes_file, 
                cohort = cohort,
                save_directory = save_directory)
```


### Estimate geometric mean concentrations of the metabolites
```{r singlePipeline}
result <- readNHANES(codes_file = codes_file, 
                       data_path = data_path,  
                       cohort = cohort,
                       save_directory = paste(save_directory, "/", cohort, sep = ""))
  
# Update the previous result based on closer examination of the metabolite concentration distributions 
result$Measured <- examine_error(Measured = result$Measured, 
                                 codes_file = codes_file, 
                                 data_path = data_path, 
                                 save_directory = paste(save_directory, "/", cohort, sep = ""))
  
# Run the Bayesian model
subpops <- "0 - 5"
output <- fitOnlyP(SUBPOP = subpops, Measured = result$Measured, mapping = result$mapping, 
                   pred.data = result$pred.data, quick = FALSE, cores = 3,
                   save_directory = paste(save_directory, "/", cohort, sep = ""))
  
# Convert to final units and check convergence
final <- fromMolar(SUBPOP = subpops, Measured = output$Measured, pred.data = output$pred.dat, 
                            nhanesdata = output$nhanesdata, out.coda3R = output$out.coda3R, doplot = TRUE,
                            save_directory = paste(save_directory, "/", cohort, sep = ""))
```


## Generate a simple plot !!! Need to update this or plot_subpops
```{r simplePlot}
plot_info <- plot_subpops(path_to_lPs = paste(save_directory, "/", cohort, sep = ""), 
                          SUBPOP = subpops, print_plot = TRUE)
```


# Looping Over Multiple Cohorts
Obtain exposure estimates for multiple consecutive years of the NHANES survey for a subset of chemicals to
observe changes in exposure over time and differences between population groups.

## Prepare for session
### Load the relevant librarires
```{r setup2}
library(bayesmarker)
```

### Assign needed function inputs 
The codes_file has specific formatting (see the example file included in this package or the manuscript supplemental data).
Examine exposure changes for four parabens.
```{r inputs2}
cohort <- c("05-06", "07-08", "09-10", "11-12", "13-14", "15-16")
subset <- c("Ethyl paraben", "n-propyl paraben", "Butyl paraben", "Methyl paraben")
subset.type <- "Name"
codes_file <- "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx"
save_directory <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
data_path <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
```

## Use the bayesmarker package 
### Obtain and save the raw data
First get all the needed data from the CDC NHANES website. This call will result in a folder for each cohort 
(name in the long form, for example "2015-2016") at the location "save_directory/rawData/".
The directory will include all needed files corresponding to the table in the codes file after subsetting by chemical
and cohorts.
```{r parabenCohorts}
get_NHANES_data(codes_file = codes_file, 
                phases = cohort,
                save_directory = save_directory)
```


### Estimate geometric mean concentrations of the metabolites
Obtain estimates per cohort using a loop over the main pipeline.
```{r indivParabens}
for (i in 1:length(cohort)){
  
  print(paste("Starting computations for cohort", cohort[i], sep = ""))
  
  # Read NHANES
  print("Starting readNHANES function")
  result <- readNHANES(codes_file = codes_file, 
                       data_path = data_path,  
                       cohort = cohort[i], subset = subset, subset.type = subset.type,
                       save_directory = paste(save_directory, "/", cohort[i], sep = ""))
  
  # Add needed columns to Measured
  print("Starting examine_error function")
  result$Measured <- examine_error(Measured = result$Measured, 
                                   codes_file = codes_file, 
                                   data_path = data_path, 
                                   save_directory = paste(save_directory, "/", cohort[i], sep = ""))
  
  # Run the model
  print("Starting fitOnlyP function")
  subpops <- unique(result$Measured$subpop)
  output <- c()
  # There are 10 population groups, so this step is computationally intensive. If a speedup is desired,
  # can run this in parallel using either foreach or mclapply.
  output <- list()
  for (j in 1:length(subpops)){
    output[[j]] <- fitOnlyP(SUBPOP = subpops[j], Measured = result$Measured, mapping = result$mapping, 
                       pred.data = result$pred.data, quick = FALSE, cores = 3,
                       save_directory = paste(save_directory, "/", cohort[i], sep = ""))
  }
  
  # Convert to final units and check convergence
  print("Starting fromMolar function")
  final <- lapply(1:length(subpops),
                  FUN = function(j) {
                    fromMolar(SUBPOP = subpops[j], Measured = output[[j]]$Measured, pred.data = output[[j]]$pred.dat, 
                              nhanesdata = output[[j]]$nhanesdata, out.coda3R = output[[j]]$out.coda3R, doplot = TRUE,
                              save_directory = paste(save_directory, "/", cohort[i], sep = ""))
                  })
}
```

## Retrieve results and generate time series plots
The MCMC chains are named in the form "save_directory/cohort/lPsamps-gm_kg_day_subpop_year-month-day.RData".



# Combining Cohorts
Providing multiple cohorts to the readNHANES function will result in combination of the survey data from 
those cohorts. This increases the sample size, leading to greater statistical power, which may result
in reduced uncertainty in the final exposure estimates. This section compares phthalte exposures to all 
individuals in the 2000s compared to the 2010s.

## Prepare for session
### Load the relevant librarires
```{r setup3}
library(bayesmarker)
library(readxl)
library(xlsx)
```

### Assign needed function inputs 
The codes_file has specific formatting (see the example file included in this package or the manuscript supplemental data).
Examine exposure changes for phthalates.
```{r inputs3}
samples <- list()
samples[[1]] <- c("01-02", "03-04", "05-06", "07-08", "09-10")
samples[[2]] <- c("11-12", "13-14", "15-16")
names(samples) <- c("2000s", "2010s")
save_directory <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
data_path <- "/share/home3/zstanfie/MetaboliteInference/testingAllPhases"
```

### Modify the codes file
To do a comparison across decades, a chemical needs to be measured in at least one cohort of each decade.
Then we reduce metabolites to just phthalates.
```{r modCodesFile}
# Ensure data for each decade
convtbl <- as.data.frame(read_excel("/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 1))
convtbl <- convtbl[!convtbl$recent_sample == "99-00", ] # Not in either decade
chems <- unique(convtbl$CAS)
indkeep <- logical(length = length(chems))
for (i in 1:length(chems)){
  presPhase <- convtbl$recent_sample[convtbl$CAS == chems[i]]
  print(chems[i])
  print(presPhase)
  if (length(intersect(presPhase, samples[[1]])) > 0 & length(intersect(presPhase, samples[[2]])) > 0){
    indkeep[i] <- TRUE
    print("keep")
  } else {
    indkeep[i] <- FALSE
    print("don't keep")
  }
}
convtbl <- convtbl[convtbl$CAS %in% chems[indkeep],]
# Remove non-phthalates
convtbl <- convtbl[grep("phthalate", convtbl$Name),]
# Update the weights table to match
keepFile <- paste(unique(convtbl$NHANESfile), ".XPT", sep = "")
wtvars <- as.data.frame(read_excel("/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 2))
wtvars <- wtvars[wtvars$file %in% keepFile,]
# Write Files
map <- as.data.frame(read_excel("/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 3))
write.xlsx(convtbl, file = "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_forDecadeComp.xlsx", sheetName = "codes")
write.xlsx(wtvars, file = "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_forDecadeComp.xlsx", 
           sheetName = "weights", append = TRUE)
write.xlsx(map, file = "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_forDecadeComp.xlsx", 
           sheetName = "map", append = TRUE)
codes_file <- "/share/home3/zstanfie/MetaboliteInference/NHANEScodes_ZS_forDecadeComp.xlsx"
```

## Use the bayesmarker package 
### Obtain and save the raw data
```{r phthalateDecadeData}
get_NHANES_data(codes_file = codes_file, 
                cohort = cohort,
                save_directory = save_directory)
```

### Loop over decades and run the main pipeline
```{r decadeMain}
for (i in 1:length(samples)){
  
  print(paste("Starting computations for ", names(samples)[i], sep = ""))
  
  # Read NHANES
  #print("Starting readNHANES function")
  result <- readNHANES(codes_file = codes_file, 
                       data_path = data_path,  
                       cohort = unname(samples[[i]]),
                       save_directory = paste(save_directory, "/", names(samples)[i], sep = ""))
  
  # Add needed columns to Measured
  print("Starting examine_error function")
  result$Measured <- examine_error(Measured = result$Measured, 
                                   codes_file = codes_file, 
                                   data_path = data_path, 
                                   save_directory = paste(save_directory, "/", names(samples)[i], sep = ""))
  
  # Run the Bayesian inference.
  print("Starting fitOnlyP function")
  subpops <- "Total"
  output <- fitOnlyP(SUBPOP = subpops, Measured = result$Measured, mapping = result$mapping, 
                       pred.data = result$pred.data, quick = FALSE, cores = 3,
                       save_directory = paste(save_directory, "/", names(samples)[i], sep = ""))

  
  # Convert to final units and check convergence
  print("Starting fromMolar function")
  final <- fromMolar(SUBPOP = subpops, Measured = output$Measured, pred.data = output$pred.dat, 
                              nhanesdata = output$nhanesdata, out.coda3R = output$out.coda3R, doplot = TRUE,
                              codes_file = codes_file,
                              save_directory = paste(save_directory, "/", names(samples)[i], sep = ""))
}
```

## Make a plot

