---
title: "Visualize Metabolite Map:  Figure 2"
author: "Zachary Stanfield"
date: "January 4, 2022"
output: html_document
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

## Prepare for session
### Load relevant libraries
```{r setup}
library(tidyverse)
library(igraph)
library(gdata)
library(RColorBrewer)
```

### Load the data
```{r mapData}
map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", as.is = TRUE, sheet = 3)
```

## Prepare the network

Want to create an edge list with columns:  parent (source) and metabolite (destination) 
with corresponding data frame with node type information

### Edge list
```{r edgeList}
data <- map[,colnames(map) %in% c("Name", "Name.1")]
colnames(data) <- c("source", "destination")
sources <- data %>%
  distinct(source) %>%
  rename(label = source)
sources$type = rep("Parent chemical", dim(sources)[1])
destinations <- data %>%
  distinct(destination) %>%
  rename(label = destination)
destinations$type = rep("Chemical measured in urine", dim(destinations)[1])
```

### Nodes table
We have 3 node types:  1. exclusively parent, 2. exclusively metabolite, and 3. chemical directly measured
in urine (could have both parents and metabolites).
```{r nodes}
nodes <- full_join(sources, destinations, by = "label")
indP <- unlist(apply(nodes, 1, function(x) x[2] == "Parent chemical"))
indM <- unlist(apply(nodes, 1, function(x) x[3] == "Chemical measured in urine"))
indB <- indP & indM
parents <- which(indP & is.na(indM))
metabs <- which(indM & is.na(indP))
both <- which(!is.na(indB))
nodes$type = NA
nodes$type[parents] <- "Parent chemical"
nodes$type[metabs] <- "Chemical measured in urine"
nodes$type[both] <- "Directly measured"
nodes <- nodes %>% rowid_to_column("id")
nodes <- nodes[,c(1,2,5)]
```

## Make final edge list
Add all the information together to a data frame that will be used for plotting
```{r edges}
per_route <- data %>%
  group_by(source, destination) %>%
  summarise(weight = n()) %>%
  ungroup()
edges <- per_route %>%
  left_join(nodes, by = c("source" = "label")) %>%
  rename(from = id)
edges <- edges %>%
  left_join(nodes, by = c("destination" = "label")) %>%
  rename(to = id)
edges <- select(edges, from, to, weight)
# Remove self loops
ind <- edges$from == edges$to
edges <- edges[!ind,]
# Final graph
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)
coul <- rev(brewer.pal(3, "Set1"))
my_color <- coul[as.numeric(as.factor(V(g)$type))]
```

## Make the plot
```{r Fig2}
#pdf(file = "Figure2.pdf", width = 10, height = 10)
plot(g, vertex.color = my_color, layout = layout_with_kk, edge.arrow.size = 0.3,
      vertex.label.cex = 0.5, vertex.label.dist = 0.25, vertex.size = 1.5, vertex.label.color = "black") #, vertex.label = NA
#dev.off()

```


