---
title: "Calculate Bioactivity:Exposure Ratios for Parent Chemicals of NHANES metabolites:  Fig. 5"
author: "Zachary Stanfield"
date: "January 6, 2022"
output: html_document
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This file uses the inferred exposures for parent chemicals of NHANES biomarkers that was obtained using
the bayesmarker package along with the R package HTTK and estimated rat LD50 values (via OPERA) to 
obtain bioactivity exposure rations (BERs), which is a measure of risk.

## Prepare for session
### Load relevant libraries
```{r setup}
library(httk)
library(gdata)
library(ggplot2)
library(reshape2)
library(colorRamps)
```

### Declare some needed data objects
```{r dataObjs}
# Reference for population groups
subpops <- c("Total", "Male", "Female", "0 - 5", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
             "ReproAgeFemale", "BMI <= 30", "BMI > 30")
names(subpops) <- c("Total", "Male", "Female", "0 - 5", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
                    "ReproAgeFemale", "BMI_le_30", "BMI_gt_30")
# Cohorts
phases <- c("99-00", "01-02", "03-04", "05-06", "07-08", "09-10", "11-12", "13-14", "15-16")
path_to_lPs <- "testingAllPhases"
map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", as.is = TRUE, sheet = 3)
```

## Obtain desired data from saved output of bayesmarker runs  
Select the most recent cohort for each parent chemical from the full set of individual cohort runs.
If this data is not already available, users can alter their input file to include just the latest
cohort row for each individual biomarker before working through the bayesmarker pipeline (see the 
vignette calculateExposureInferences for the pipeline).  

```{r readLPs}
dta.1 <- list()
dta.2 <- list()
for (j in 1:length(phases)){
  
  # Obtain file names for lPsamps
  if (is.null(path_to)){
    files <- list.files(path = ".", pattern = "lPsamps-gm_kg_day")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  } else {
    path <- file.path(".", path_to, phases[j])
    files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Load each subpop file, get 95% CI, convert units, and save
  for (i in 1:length(subpops)){
    if (length(grep(paste("_", subpops[i], sep = ""), files)) == 0){
      # For cases like "0 - 5" supbop not in other phases.
      # Ok to call dta.1 here b/c 0-5 is not 1st in subpops
      dta <- NA
    } else {
      load(files[grep(paste("_", subpops[i], sep = ""), files)])
      samps <- exp(as.matrix(lPsampsgm))
      dta <- as.data.frame(samps)
    }
    dta.1[[i]] <- dta
    names(dta.1)[i] <- subpops[i]
  }
  dta.2[[j]] <- dta.1
  names(dta.2)[j] <- phases[j]
}

# Get the most recent estimate for each chemical. Read in previous most-recent result for easier filtering
load("ParentExpEsts_MostRecent_withCIquant_2021-09-24.RData")
chems <- final[,3]
ExpByChem <- list() 
for (i in 1:length(chems)){
  cohort <- final[final[,3] == chems[i],37] 
  for (j in 1:length(subpops)){
    if (j == 1){
      ExpByChem[[i]] <- dta.2[[cohort]][[subpops[j]]][,chems[i]]
    } else {
      if (is.null(dim(dta.2[[cohort]][[subpops[j]]]))){  # No data for this population in this cohort
        ExpByChem[[i]] <- cbind(ExpByChem[[i]], as.data.frame(matrix(data = NA, nrow = 1000, ncol = 1)))
      } else {
        if (!chems[i] %in% colnames(dta.2[[cohort]][[subpops[j]]])){  # Chemical not measured for this population in this cohort
          ExpByChem[[i]] <- cbind(ExpByChem[[i]], as.data.frame(matrix(data = NA, nrow = 1000, ncol = 1)))
        } else {
          ExpByChem[[i]] <- cbind(ExpByChem[[i]], dta.2[[cohort]][[subpops[j]]][,chems[i]])
        }
      }
    }
  }
  ExpByChem[[i]] <- as.data.frame(ExpByChem[[i]])
  colnames(ExpByChem[[i]]) <- subpops
}
names(ExpByChem) <- chems

```


## Obtain LD50 values
If this is first runthrough of this file, need to save chemical identifiers and run
through the OPERA software to obtain LD50 values using CATMoS

```{r forOPERA}
# Unique chemical identifiers to obtain CATMoS predictions
ind <- match(names(ExpByChem), map$CAS)
chemDTXSIDs <- unique(map$DTXSID[ind])
write.table(chemDTXSIDs, file = "parentChemDTXSIDs.txt", col.names = FALSE, row.names = FALSE, quote = FALSE)
# Run these chemicals through OPERA

# Load those results (CATMoS data)
testpreds <- read.xls("parentChemDTXSIDS_OPERAout_CATMoS.xlsx")
testpreds <- testpreds[!is.na(testpreds$CATMoS_VT_pred), 1:12]

# Apply uncertainty factor (Venman and Flaga 1985)
testpreds$CATMoS_LD50_pred <- 0.0001 * testpreds$CATMoS_LD50_pred
```

## Use HTTK to calculate the human equivalent dose based on the rat LD50
```{r calcHED}
# For which chemicals does hhtk have all the information we need?
z <- subset(get_cheminfo(info = "all", species = "Human", model = "pbtk"), DTXSID %in% testpreds$MoleculeID)

# Use the calc_tkstats function to get mean concentration from CATMoS value.
# Already in correct units of mg/kg
tmp_risk <- testpreds[testpreds$MoleculeID %in% z$DTXSID,]
ind <- match(tmp_risk$MoleculeID, map$DTXSID)
tmp_risk$CASRN <- map$CAS[ind]
redmap <- map[match(tmp_risk$CASRN, map$CAS), c("Name", "CAS", "DTXSID", "MW")]
risk <- c()
for (i in 1:dim(tmp_risk)[1]){
  dose_value <- tmp_risk$CATMoS_LD50_pred[i]
  risk[i] <- calc_tkstats(chem.cas = tmp_risk$CASRN[i], route = "oral", species = "Human", days = 1, daily.dose = 1, 
                          dose = dose_value, doses.per.day = 1, output.units = "mg/L", concentration = "plasma", stats = "mean",
                          tissue = "plasma", model = "pbtk", default.to.human = TRUE)
  
}
names(risk) <- tmp_risk$CASRN
```


## Use the HTTK calc_mc_css function to get steady state concentration (is 1 mg/kg/day) from our inferred
## exposure values

```{r calcExp}
y <- subset(get_cheminfo(info = "all", species = "Human", model = "pbtk"), DTXSID %in% testpreds$MoleculeID)

# Need to get risk for each subpopulation
tmp_exp_risk <- testpreds[testpreds$MoleculeID %in% y$DTXSID,]

ind <- match(tmp_exp_risk$MoleculeID, map$DTXSID)
tmp_exp_risk$CASRN <- map$CAS[ind]

# Step through all chemicals to get the css value
ssconc <- c()
exp_risk <- list()
for (i in 1:dim(tmp_exp_risk)[1]){
  ind <- which(names(ExpByChem) == tmp_exp_risk$CASRN[i])
  tmp <- ExpByChem[[ind]]
  out <- calc_mc_css(chem.cas = tmp_exp_risk$CASRN[i], samples = 1000, which.quantile = 0.50, species = "Human",
                          model = "pbtk", output.units = "mg/L")  
  ssconc <- c(ssconc, out)
  exp_risk[[i]] <- out * tmp
}
names(exp_risk) <- tmp_exp_risk$CASRN
```


## Obtain the BERs using the transformed exposure values and human equivalent dose
```{r calcBER}
ratio <- list()
for (i in 1:length(exp_risk)){
  ratio[[i]] <- risk[i]/exp_risk[[i]]  
}
names(ratio) <- names(exp_risk)
```


## Set up DF and plot: step through all chemicals and all subpops for each chemical
```{r plotSetUp}
# Get CIs for Total and just median for others
df <- as.data.frame(matrix(data = NA, nrow = length(ratio), ncol = length(subpops) + 2, 
                           dimnames = list(names(ratio), c(subpops, "lowCI", "upCI"))))
for (i in 1:length(ratio)){
  for (j in 1:length(ratio[[i]])){
    if (j == 1){
      tmp <- quantile(ratio[[i]][,j], probs = c(0.5, 0.025, 0.975), na.rm = TRUE)
      df[i,subpops[j]] <- tmp[1] 
      df[i, c("lowCI", "upCI")] <- tmp[2:3]
    } else {
    df[i,subpops[j]] <- quantile(ratio[[i]][,j], probs = 0.5, na.rm = TRUE)
    }
  }
}

# Add preferred name and DTXSID
indP <- match(row.names(df), map$CAS)
df$Chemical <- map$Name[indP]
df$DTXSID <- map$DTXSID[indP]

# Romove DNOP as it is a major outlier 
df <- df[!df$Chemical == "Di-n-octyl phthalate",]
# And Ethylbenzene has too many values below the LOD
df <- df[!df$Chemical == "Ethylbenzene",]

# Add chemical class
map <- map[map$CAS %in% row.names(df),]
map <- map[,c(2,3,4,7,8,9)]
map$CAS.1 <- gsub(" ", "", map$CAS.1)
classTable <- read.xls("allPhasesAllChems_ChemClass.xlsx", sheet = 1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])

# Shorten the class names for a more concise plot
map$ChemClass <- as.character(map$ChemClass)
map$ChemClass[map$ChemClass == "Carbamate Pesticide Metabolites"] <- "Carbamate Pesticides"
map$ChemClass[map$ChemClass == "Flame Retardant Metabolite"] <- "Flame Retardant"
map$ChemClass[map$ChemClass == "Fungicides and Metabolites"] <- "Fungicides"
map$ChemClass[map$ChemClass == "Herbicides and Metabolites"] <- "Herbicides"
map$ChemClass[map$ChemClass == "Insect Repellent and Metabolites"] <- "Insect Repellent"
map$ChemClass[map$ChemClass == "Organochlorine Pesticide Metabolites"] <- "Organochlorine Pesticides"
map$ChemClass[map$ChemClass == "Personal Care and Consumer Product Chemicals"] <- "Personal Care/Consumer Product"
map$ChemClass[map$ChemClass == "Phthalate and Phthalate Alternative Metabolites"] <- "Phthalates"
map$ChemClass[map$ChemClass == "Phytoestrogens and Metabolites"] <- "Phytoestrogens"
map$ChemClass[map$ChemClass == "Polycyclic Aromatic Hydrocarbon Metabolites"] <- "Polycyclic Aromatic Hydrocarbons"
map$ChemClass[map$ChemClass == "Pyrethroid Metabolites"] <- "Pyrethroids"
map$ChemClass[map$ChemClass == "Volatile Organic Compound Metabolites"] <- "Volatile Organic Compounds"

# Fix up factor levels
tmp <- as.character(map$ChemClass[match(row.names(df), map$CAS)])
tmp[is.na(tmp)] <- "Unknown"
df$ChemicalClass <- tmp
df$Chemical <- factor(df$Chemical, levels = as.character(df$Chemical))
df$ChemicalClass <- factor(df$ChemicalClass, levels = sort(unique(df$ChemicalClass)))
df <- melt(df, id.vars = c("Chemical", "ChemicalClass", "upCI", "lowCI"))
colnames(df) <- c("Chemical", "ChemicalClass", "upCI", "lowCI", "Group", "BER")

# Clean up
df <- df[!df$Group == "DTXSID",]
df$BER <- as.numeric(df$BER)

# Sort by lowest BER
ind <- order(df$BER[df$Group == "Total"], decreasing = FALSE)
df$Chemical <- ordered(df$Chemical, levels = levels(df$Chemical)[ind])
```

## Make final plot
```{r BERplot}
pdf(file = paste("BERplotForParents_subpops_withCIs_0.0001UF_primaryCols_", 
                 format(Sys.time(), "%Y-%m-%d"), ".pdf", sep = ""), width = 17.5, height = 9.5)
ggplot(df, aes(x = Chemical, y = BER, color = ChemicalClass, group = Group, shape = Group)) + 
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lowCI, ymax = upCI), width = 0.2, alpha = 0.8) +
  scale_y_log10(limits = c(1, 1e+12)) +
  ylab("BER") +
  ggtitle("Bioactivity:Exposure Ratio (BER) for Parent Chemicals of NHANES Metabolites") +
  theme_set(theme_gray(base_size = 17)) +
  theme(axis.text.x=element_text(angle=50, hjust=1, size = 13)) +
  theme(axis.title.x = element_blank()) +
  scale_color_manual(values = primary.colors(n = 1 + length(levels(df$ChemicalClass)))) +
  #scale_color_manual(values = rainbow(n = length(levels(df$ChemicalClass)))) +
  scale_shape_manual(name = "Group", 
                     labels = c("Total", "Males", "Females", "Age.0.5", "Age.6.11", "Age.12.19", "Age.20.65", 
                                "Age.GT65", "ReproAgeFemale", "BMI <= 30", "BMI > 30"),
                     values = c(0,1,2,3,4,5,8,11,15,16,18)) 
dev.off()
```

# Save supplemental data file
```{r suppData}
ind <- match(df$Chemical, map$Name)
df_out <- cbind("DTXSID" = map$DTXSID[ind], "CAS" = map$CAS[ind], df)
df_out <- df_out[,c(1:4,7,8,6,5)]
write.table(df_out, file = "bayesmarkerPaper_Fig5_data.txt", sep = "\t", row.names = FALSE)
```


















