---
title: "Compare Child Populations with All Individuals (Fig. 6)"
author: "Zachary Stanfield"
date: "January 11, 2022"
output: html_document
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This file uses the inferred exposures for parent chemicals of NHANES biomarkers that was obtained using
the bayesmarker package to compare exposures between young children (0 - 5 years) and children (6 - 11 
years) to all survey participants using the 2015-2016 cohort data.

## Prepare for session
### Load relevant libraries
```{r setup}
library("coda")
library(reshape2)
library(ggplot2)
library(mratios)
```

### Declare some needed data objects
```{r dataObjs}
subpops <- c("0 - 5", "6 - 11 years", "Total")
names(subpops) <- c("0 - 5", "6 - 11 years", "Total")

path <- "./testingAllPhases/15-16"

map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", sheet = 3)
classTable <- read.xls("allPhasesAllChems_ChemClass.xlsx", sheet = 1)
```

## Collect the data
If not already generated, run the main vignette for the 2015-2016 cohort for all metabolites and the 3 above "subpops"
```{r collectData}
alldata <- list()
for (i in 1:length(subpops)){
  file <- paste(path, "/", list.files(path = path, pattern = paste("lPsamps-gm_kg_day_", subpops[i], sep = "")), sep = "")
  file <- file[grep("2021-05-11", file)]
  load(file)
  samps <- as.matrix(lPsampsgm)
  alldata[[i]] <- exp(samps)
  names(alldata)[i] <- subpops[i]
}
```

## Calculate the fold changes
```{r calcFC}
df1 <- data.frame("Chemical" = colnames(alldata[[1]]), "FC" = NA, "lowCI" = NA, "upCI" = NA, "p.value" = NA)
df2 <- df1
numsamps <- dim(alldata[[1]])[1]
for (i in 1:dim(df1)[1]){
  temp <- ttestratio(alldata$`0 - 5`[sample.int(numsamps, 2000),i], alldata$Total[sample.int(numsamps, 2000),i], 
                     alternative = "two.sided", conf.level = 0.95)
  df1[i,c("FC", "lowCI", "upCI", "p.value")] <- c(temp$estimate["x/y"], temp$conf.int[1:2], temp$p.value)
  temp <- ttestratio(alldata$`6 - 11 years`[sample.int(numsamps, 2000),i], alldata$Total[sample.int(numsamps, 2000),i], 
                     alternative = "two.sided", conf.level = 0.95)
  df2[i,c("FC", "lowCI", "upCI", "p.value")] <- c(temp$estimate["x/y"], temp$conf.int[1:2], temp$p.value)
}
```


## Prepare the data for the plot
```{r prepData}
# One chemical has a number of outliers, remove this
df1 <- df1[-grep("91-20-3", df1$Chemical),]
df2 <- df2[-grep("91-20-3", df2$Chemical),]

# Record order for largest FC for 0-5:Total
sort_ind <- order(df1$FC, decreasing = TRUE)

# Prepare for final data frame
df1 <- cbind(df1, "Comparison" = "0-5:Total")
df2 <- cbind(df2, "Comparison" = "6-11:Total")
df <- rbind(df1, df2)

# Add chemical class
# Propagate reported (by NHANES) metabolite class to the parent chemicals using
#the metabolite map
map <- map[map$CAS %in% df1$Chemical,]  
map$ChemClass <- classTable$Column4[match(map$CAS.1, classTable$Column1)]

# Shorten the class names for a more concise plot
map$ChemClass <- as.character(map$ChemClass)
map$ChemClass[map$ChemClass == "Carbamate Pesticide Metabolites"] <- "Carbamate Pesticides"
map$ChemClass[map$ChemClass == "Flame Retardant Metabolite"] <- "Flame Retardant"
map$ChemClass[map$ChemClass == "Fungicides and Metabolites"] <- "Fungicides"
map$ChemClass[map$ChemClass == "Herbicides and Metabolites"] <- "Herbicides"
map$ChemClass[map$ChemClass == "Insect Repellent and Metabolites"] <- "Insect Repellent"
map$ChemClass[map$ChemClass == "Organochlorine Pesticide Metabolites"] <- "Organochlorine Pesticides"
map$ChemClass[map$ChemClass == "Personal Care and Consumer Product Chemicals"] <- "Personal Care/Consumer Product"
map$ChemClass[map$ChemClass == "Phthalate and Phthalate Alternative Metabolites"] <- "Phthalates"
map$ChemClass[map$ChemClass == "Phytoestrogens and Metabolites"] <- "Phytoestrogens"
map$ChemClass[map$ChemClass == "Polycyclic Aromatic Hydrocarbon Metabolites"] <- "Polycyclic Aromatic Hydrocarbons"
map$ChemClass[map$ChemClass == "Pyrethroid Metabolites"] <- "Pyrethroids"
map$ChemClass[map$ChemClass == "Volatile Organic Compound Metabolites"] <- "Volatile Organic Compounds"

# If a parent chemical has metabolites from more than one class, label NA (unknown)
df$ChemicalClass <- NA
tmp <- list()
for (i in 1:dim(df)[1]){
  tmp[[i]] <- unique(as.character(map$ChemClass[as.character(map$CAS) == as.character(df$Chemical)[i]]))
  if (length(tmp[[i]]) == 1){
    df$ChemicalClass[i] <- tmp[[i]]  
  }
}

# Add chemical name
df$Name <- map$Name[match(df$Chemical, map$CAS)]
df$Name <- droplevels(df$Name)
df$Name <- ordered(df$Name, levels = as.character(df$Name[df$Comparison == "0-5:Total"])[sort_ind])

# Offer space for wrapping of facet label
df$ChemicalClass[df$ChemicalClass == "Perchlorate and Other Anions"] <- "Per- chlorate and Other Anions"
```


# Make the final plot and save the data for supplement
```{r plot}
pdf(file = "FCs_DiffTotalvs0to5vs6to11_withCIs_2021-10-14.pdf", width = 16, height = 10)
ggplot(df, aes(Name, FC, fill = Comparison)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = lowCI, ymax = upCI), width = 0.2, position = position_dodge(0.9)) +
  scale_fill_manual(values = c("#E69F00", "#999999")) +
  facet_grid(~Class, scales = "free_x", space = "free_x", labeller = labeller(Class = label_wrap_gen(10))) +
  ylab("Fold Change") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  ggtitle("Exposure Fold Change Comparison Between the Total Population and Children aged 0-5 and 6-11") +
  theme(text = element_text(size = 17)) +
  theme(axis.title.x = element_blank()) +
  theme(axis.text.x = element_text(size = 13, angle = 60, hjust = 1)) +
  theme(strip.text.x = element_text(size = 12)) +
  theme(panel.spacing = unit(0.5, "lines"))
dev.off()

# Save data for manuscript
ind <- match(df$Chemical, map$CAS)
df <- cbind("DTXSID" = map$DTXSID[ind], df)
df <- df[,c(1,2,9,8,7,3:5)]
colnames(df)[2] <- "CAS"
write.table(df, file = "bayesmarkerPaper_Fig6_data.txt", sep = "\t", row.names = FALSE)
```
















