---
title: "Compare SEEM Predictions to Bayesmarker Inferences (Fig. 7)"
author: "Zachary Stanfield"
date: "January 11, 2022"
output: html_document
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This file compares the infered exposure estimates obtained from bayesmarker with the SEEM predictions of exposure
for "new" chemicals (those chemicals not in the SEEM calibration set) for the 4 cohorts since the first iteration
of SEEM (2009-2010, 2011-2012, 2013-2014, and 2015-2016).

## Prepare for session
### Load relevant libraries
```{r setup}
library(gdata)
library(ggplot2)
```

### Declare some needed data objects
```{r dataObjs}
# Read SEEM3 predictions
load("chem.preds-2018-11-28.RData")
colnames(chem.preds)

# Load old estimates to drop chemicals that were in the original training data
oldEsts <- read.xls("Wambaugh_et_al_2014_urine.xlsx", pattern = "Compound")

# Reference for population groups
subpops <- c("Total", "Male", "Female", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
             "ReproAgeFemale", "BMI <= 30", "BMI > 30")
names(subpops) <- c("Total", "Male", "Female", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
                    "ReproAgeFemale", "BMI_le_30", "BMI_gt_30")

# Subpops the user wants plotted
SUBPOP <- "Total"
if (SUBPOP != "all"){
  ind <- SUBPOP %in% subpops
  subpops <- subpops[ind]
}

phases <- c("09-10", "11-12", "13-14", "15-16")

path_to_lPs <- "testingAllPhases"
```

## Obtain the bayesmarker infered exposures for the relevant population and cohorts
If this data has not yet been calculated, users can run the main vignette for the 4 above cohots and the total
population for either all chemicals or the non-calibration set of chemicals (compare the set of chemicals of 
each cohort with the ones in oldEsts and chem.preds to just get inferences for the desired chemicals)
```{r getData}
dta.2 <- c()
dta.2.low <- c()
dta.2.up <- c()
for (j in 1:length(phases)){
  
  # Obtain file names for lPsamps
  if (is.null(path_to_lPs)){
    files <- list.files(path = ".", pattern = "lPsamps-gm_kg_day")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  } else {
    path <- file.path(".", path_to_lPs, phases[j])
    files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Read in files and build data.frame
  load(files[grep(paste("_", subpops, sep = ""), files)])
  samps <- as.matrix(lPsampsgm)
  
  # Having all NAs for a column causes an error
  rmind <- unlist(apply(samps, 2, function(x) sum(is.na(x)) == length(x)))
  samps <- samps[,!rmind]
  result.med <- apply(samps, 2, mean)
  result.bar <- mean(apply(samps, 1, mean))
  
  # Get 95th Percentiles
  result.CI <- t(apply(samps, 2, function(x) quantile(x, probs = c(0.0275, 0.975))))
  dta <- data.frame(colnames(samps), exp(result.med))
  dta.low <- data.frame(colnames(samps), exp(result.CI[,1]))
  dta.up <- data.frame(colnames(samps), exp(result.CI[,2]))
  colnames(dta) <- c("CAS", subpops)
  colnames(dta.low) <- c("CAS", paste(subpops, "_low95", sep = ""))
  colnames(dta.up) <- c("CAS", paste(subpops, "_up95", sep = ""))

  dta.2 <- rbind(cbind(dta, "Phase" = rep(phases[j], dim(dta)[1])), dta.2)
  dta.2.low <- rbind(cbind(dta.low, "Phase" = rep(phases[j], dim(dta.low)[1])), dta.2.low)
  dta.2.up <- rbind(cbind(dta.up, "Phase" = rep(phases[j], dim(dta.up)[1])), dta.2.up)
}
```

## Remove rows with the chemicals that were in the calibration data
```{r remCalib}
dta.2 <- dta.2[!dta.2$CAS %in% oldEsts$CASRN,]
dta.2.low <- dta.2.low[!dta.2.low$CAS %in% oldEsts$CASRN,]
dta.2.up <- dta.2.up[!dta.2.up$CAS %in% oldEsts$CASRN,]
```

## Reduce SEEM and our data frames to this set of chemicals
```{r redChems}
shared <- (intersect(unique(dta.2$CAS), chem.preds$CAS)) 
SEEM <- chem.preds[chem.preds$CAS %in% shared, c("dsstox_substance_id", "CAS", "Substance_Name", "seem3", "seem3.l95", "seem3.u95")]
dta.2 <- dta.2[dta.2$CAS %in% shared,]
dta.2.low <- dta.2.low[dta.2.low$CAS %in% shared,]
dta.2.up <- dta.2.up[dta.2.up$CAS %in% shared,]
```

## Add chemical class to see if we predict better for some classes than others
```{r addClass}
map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", as.is = TRUE, sheet = 3)
map <- map[,c(2,3,4,7,8,9)]
classTable <- read.xls("allPhasesAllChems_ChemClass.xlsx", sheet = 1)
map$CAS.1 <- gsub(" ", "", map$CAS.1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])

# Shorten the class names for concise plotting
map$ChemClass <- as.character(map$ChemClass)
map$ChemClass[map$ChemClass == "Carbamate Pesticide Metabolites"] <- "Carbamate Pesticides"
map$ChemClass[map$ChemClass == "Flame Retardant Metabolite"] <- "Flame Retardant"
map$ChemClass[map$ChemClass == "Fungicides and Metabolites"] <- "Fungicides"
map$ChemClass[map$ChemClass == "Herbicides and Metabolites"] <- "Herbicides"
map$ChemClass[map$ChemClass == "Insect Repellent and Metabolites"] <- "Insect Repellent"
map$ChemClass[map$ChemClass == "Organochlorine Pesticide Metabolites"] <- "Organochlorine Pesticides"
map$ChemClass[map$ChemClass == "Personal Care and Consumer Product Chemicals"] <- "Personal Care/Consumer Product"
map$ChemClass[map$ChemClass == "Phthalate and Phthalate Alternative Metabolites"] <- "Phthalates"
map$ChemClass[map$ChemClass == "Phytoestrogens and Metabolites"] <- "Phytoestrogens"
map$ChemClass[map$ChemClass == "Polycyclic Aromatic Hydrocarbon Metabolites"] <- "Polycyclic Aromatic Hydrocarbons"
map$ChemClass[map$ChemClass == "Pyrethroid Metabolites"] <- "Pyrethroids"
map$ChemClass[map$ChemClass == "Volatile Organic Compound Metabolites"] <- "Volatile Organic Compounds"
```


## Make a facet plot by cohort 
```{r makePlot}
indclass <- match(dta.2$CAS, map$CAS)
ind <- match(dta.2$CAS, SEEM$CAS)

df <- data.frame(y = dta.2$Total, x = SEEM$seem3[ind], ylower = dta.2.low$Total_low95, yupper = dta.2.up$Total_up95,
                 xlower = SEEM$seem3.l95[ind], xupper = SEEM$seem3.u95[ind], ChemicalClass = map$ChemClass[indclass],
                 "NHANES.cohort" = dta.2$Phase)
df$NHANES.cohort <- factor(df$NHANES.cohort, levels = phases)

pdf(file = "Comparison_NewInferencesVsSEEM_facet_2021-07-09.pdf", width = 18.5, height = 16)
ggplot(data = df, aes(x = x, y = y, ymin = ylower, ymax = yupper, xmin = xlower, xmax = xupper, color = ChemicalClass, group = NHANES.cohort)) +
  geom_point(size = 1.75) +
  geom_errorbar(width = 0) +
  geom_errorbarh(width = 0) +
  geom_abline(xintercept=0, yintercept=1, lty=3, color="darkgray") +
  scale_y_log10() +
  scale_x_log10() +
  facet_wrap(~NHANES.cohort, nrow = 2) +
  theme_set(theme_gray(base_size = 23)) +
  theme(legend.position = "bottom") +
  ylab("Updated NHANES Inferences") +
  xlab("SEEM Predictions") +
  coord_flip() + 
  ggtitle("Comparison of Exposure Intake Rates for bayesmarker and SEEM")
dev.off()
```

## Save data for supplemental table
```{r saveData}
ind2 <- match(dta.2$CAS, map$CAS)
df <- data.frame("Cohort" = dta.2$Phase, "DTXSID" = map$DTXSID[ind2], "CAS" = map$CAS[ind2], "Name" = map$Name[ind2], 
                 "ChemicalClass" = map$ChemClass[indclass], "Bayesmarker_median" = dta.2$Total, 
                 "Bayesmarker_low95CI" = dta.2.low$Total_low95, "Bayesmarker_up95CI" = dta.2.up$Total_up95,
                 "SEEM_median" = SEEM$seem3[ind], "SEEM_low95CI" = SEEM$seem3.l95[ind], "SEEM_up95CI" = SEEM$seem3.u95[ind])
df <- df[dim(df)[1]:1,]
write.table(df, file = "bayesmarkerPaper_Fig7_data.txt", sep = "\t", row.names = FALSE)
```



# Generating the R^2 supplemental table 
We take some similar steps during setup but generate r^2 values by cohort and by chemical class to better
understand how the predictions match up.  Uses the same data objects declared in the previous section.

## Obtain the bayesmarker infered exposures for the relevant population and cohorts
If this data has not yet been calculated, users can run the main vignette for the 4 above cohots and the total
population for either all chemicals or the non-calibration set of chemicals (compare the set of chemicals of 
each cohort with the ones in oldEsts and chem.preds to just get inferences for the desired chemicals)
```{r collectInfExps}
dta.2 <- list()
for (j in 1:length(phases)){
  
  # Obtain file names for lPsamps
  path <- file.path(".", path_to, phases[j])
  files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
  dates <- "2021-05-11"
  files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  
  # Load the Total file
  load(files[grep(paste("_", subpops, sep = ""), files)])
  samps <- exp(as.matrix(lPsampsgm))
  dta <- as.data.frame(samps)
  
  # Record data for this phase
  dta.2[[j]] <- dta
  names(dta.2)[j] <- phases[j]
}
# dta.2 is a list with 1 element for each of the 4 phases.  Each phase only
# has 1 table, which is the lP matrix for the total population
```

## Remove rows with the chemicals that were in the calibration data
```{r remCalib}
dta.2 <- lapply(dta.2, function(x) x[,!colnames(x) %in% as.character(oldEsts$CASRN)])
```

## Reduce SEEM and our data frames to this set of chemicals
```{r redChems}
shared <- intersect(unique(unlist(sapply(dta.2, function(x) colnames(x)))), chem.preds$CAS) 
ind <- match(shared, chem.preds$CAS)
SEEM <- chem.preds[ind, c("dsstox_substance_id", "CAS", "Substance_Name", "seem3", "seem3.l95", "seem3.u95")]
dta.2 <- lapply(dta.2, function(x) x[,colnames(x) %in% shared])
```

## Add chemical class to see if we predict better for some classes than others
```{r addClass}
map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", as.is = TRUE, sheet = 3)
map <- map[,c(2,3,4,7,8,9)]
classTable <- read.xls("allPhasesAllChems_ChemClass.xlsx", sheet = 1)
map$CAS.1 <- gsub(" ", "", map$CAS.1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])

# Shorten the class names for concise plotting
map$ChemClass <- as.character(map$ChemClass)
map$ChemClass[map$ChemClass == "Carbamate Pesticide Metabolites"] <- "Carbamate Pesticides"
map$ChemClass[map$ChemClass == "Flame Retardant Metabolite"] <- "Flame Retardant"
map$ChemClass[map$ChemClass == "Fungicides and Metabolites"] <- "Fungicides"
map$ChemClass[map$ChemClass == "Herbicides and Metabolites"] <- "Herbicides"
map$ChemClass[map$ChemClass == "Insect Repellent and Metabolites"] <- "Insect Repellent"
map$ChemClass[map$ChemClass == "Organochlorine Pesticide Metabolites"] <- "Organochlorine Pesticides"
map$ChemClass[map$ChemClass == "Personal Care and Consumer Product Chemicals"] <- "Personal Care/Consumer Product"
map$ChemClass[map$ChemClass == "Phthalate and Phthalate Alternative Metabolites"] <- "Phthalates"
map$ChemClass[map$ChemClass == "Phytoestrogens and Metabolites"] <- "Phytoestrogens"
map$ChemClass[map$ChemClass == "Polycyclic Aromatic Hydrocarbon Metabolites"] <- "Polycyclic Aromatic Hydrocarbons"
map$ChemClass[map$ChemClass == "Pyrethroid Metabolites"] <- "Pyrethroids"
map$ChemClass[map$ChemClass == "Volatile Organic Compound Metabolites"] <- "Volatile Organic Compounds"
```

## Keep a reference file for the chemical class for each cohort to help in plotting
```{r classRef}
class_ref <- list()
for (i in 1:length(dta.2)){
  phase_chems <- colnames(dta.2[[i]])
  indclass <- match(phase_chems, map$CAS)
  class_ref[[i]] <- data.frame("Chemical" = phase_chems, "Name" = map$Name[indclass], "ChemicalClass" = map$ChemClass[indclass])
  class_ref[[i]]$ChemicalClass <- factor(class_ref[[i]]$ChemicalClass, levels = c(levels(class_ref[[i]]$ChemicalClass), "Unknown"))
  class_ref[[i]]$ChemicalClass[is.na(class_ref[[i]]$ChemicalClass)] <- "Unknown"
}
names(class_ref) <- names(dta.2)
```

## Log transform the data
```{r log}
SEEM[,4:6] <- log(SEEM[,4:6])
dta.2 <- lapply(dta.2, log)
```

## Calculate R^2 values by cohort and chemical class
Also includes the option to plot these inividual data points within all cohort-class pairings
with 95% CIs, a correlation value, and associate p-value.
```{r genRsqData}
iter <- 1000
rsquar_class <- list()
rmse_class <- list()
rsquar <- as.data.frame(matrix(data = NA, nrow = iter, ncol = length(phases), dimnames = list(1:iter, phases)))
rmse <- rsquar
doPlot <- FALSE
if (doPlot){
  pdf(file = "SEEM_vs_Bayesmarker_nonTrainingChems_byChemClass_invFC.pdf", width = 9, height = 8)
}
for (j in 1:length(phases)){
  dfp <- dta.2[[j]]
  # Do a class-specific calculation as well
  class_set <- as.character(unique(class_ref[[j]]$ChemicalClass))
  rsquar_class[[j]] <- as.data.frame(matrix(data = NA, nrow = iter, ncol = length(class_set), dimnames = list(1:iter, class_set)))
  rmse_class[[j]] <- as.data.frame(matrix(data = NA, nrow = iter, ncol = length(class_set), dimnames = list(1:iter, class_set)))
  for (i in 1:iter){
    mySample <- sample.int(3000, 1000, replace = FALSE)
    for (k in 1:length(class_set)){
      class_chems <- as.character(class_ref[[j]]$Chemical[class_ref[[j]]$ChemicalClass == class_set[k]])
      dfpc <- dfp[,colnames(dfp) %in% class_chems]
      if (is.null(dim(dfpc))){
        rsquar_class[[j]][i,k] <- NA
        rmse_class[[j]][i,k] <- NA
        next
      }
      dfpc <- dfpc[mySample,]
      dfpc.med <- apply(dfpc, 2, function(x) quantile(x, probs = 0.5))
      dfpc.up <- apply(dfpc, 2, function(x) quantile(x, probs = 0.975))
      dfpc.low <- apply(dfpc, 2, function(x) quantile(x, probs = 0.0275))
      sd <- (dfpc.up - dfpc.low)
      indSEEM <- match(class_chems, SEEM$CAS)
      finalDF <- data.frame("Chemical" = class_chems, "y" = SEEM$seem3[indSEEM], "x" = dfpc.med, 
                            "xlower" = dfpc.low, "xupper" = dfpc.up, "ylower" = SEEM$seem3.l95[indSEEM],
                            "yupper" = SEEM$seem3.u95[indSEEM], "SEEMrange" = (SEEM$seem3.u95[indSEEM]/SEEM$seem3.l95[indSEEM]), 
                            "Myrange" = 1/sd^2)
      model <- lm(y ~ x, finalDF, weights = Myrange)
      
      tmp <- summary(model)
      rsquar_class[[j]][i,k] <- tmp$r.squared
      rmse_class[[j]][i,k] <- sqrt(mean(model$residuals^2))
      if (doPlot){
        g <- ggplot(data = finalDF, aes(x = x, y = y, ymin = ylower, ymax = yupper, xmin = xlower, xmax = xupper)) +
          geom_point(size = 1.75) +
          geom_errorbar(width = 0.15, size = 0.2) +
          geom_errorbarh(height = 0.15, size = 0.2) +
          geom_smooth(method = "lm") +
          geom_abline(xintercept=0, yintercept=1, lty=3, color="darkgray") +
          scale_y_log10() +
          scale_x_log10() +
          theme(legend.position = "bottom") +
          xlab("SEEM") +
          ylab("bayesmarker") +
          stat_cor(label.y.npc = "middle") + 
          ggtitle(paste("Exposure predictions: bayesmarker vs SEEM (", phases[j], ", ", class_set[k], ")", sep = "")) +
          theme(plot.title = element_text(size = 10))
        print(g)
      }
    }
    
    tmp <- summary(model)
    rsquar[i,j] <- tmp$r.squared
    rmse[i,j] <- sqrt(mean(model$residuals^2))
  }
}
dev.off()
```

## Create final data file of R^2 values
```{r RsqDataFile}
# Supplemental R^2 table: Make columns the cohorts, then 1st row is total and rest are the classes
names(rsquar_class) <- phases
supptables <- c()
for (i in 1:length(phases)){
  total <- quantile(rsquar[,phases[i]], c(0.5, 0.0275, 0.975))
  classRes <- t(apply(rsquar_class[[i]], 2, function(x) quantile(x, probs = c(0.5, 0.0275, 0.975), na.rm = TRUE)))
  temp <- rbind(total, classRes)  
  supptables <- rbind(supptables, data.frame("Cohort" = rep(phases[i], dim(temp)[1]), temp))
}
supptables$Group <- row.names(supptables)
supptables$Group[grep("total", supptables$Group)] <- "Total"
colnames(supptables) <- c("Cohort", "Median", "low95CI", "up95CI", "Group")
supptables <- supptables[,c(1,5,2,3,4)]
supptables$Group <- gsub("[0-9 ]", "", supptables$Group)
supptables[,3:5] <- round(supptables[,3:5], digits = 4)
write.table(supptables, file = "bayesmarkerPaper_Fig7_Rsquar_data.txt", sep = "\t", row.names = FALSE)
```





