---
title: "Exposure Heat Map: Log2(Fold Change) compared to the total population (Fig. 4)"
author: "Zachary Stanfield"
date: "January 6, 2022"
output: html_document
---

from "Bayesian Inference of Chemical Exposures from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This file uses the inferred exposures for parent chemicals of NHANES biomarkers that was obtained using
the bayesmarker package to generate a heatmap that shows how exposure in various population groups compares
to all survey participants.

## Prepare for session
### Load relevant libraries
```{r setup}
library(ggplot2)
library(reshape2)
library(gdata)
library(RColorBrewer)
library(colorRamps)
library(pheatmap)
```

### Declare some needed data objects
```{r dataObjs}
# Reference for population groups
subpops <- c("Total", "Male", "Female", "0 - 5", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
             "ReproAgeFemale", "BMI <= 30", "BMI > 30")
names(subpops) <- c("Total", "Male", "Female", "0 - 5", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
                    "ReproAgeFemale", "BMI_le_30", "BMI_gt_30")
# Cohorts
phases <- c("99-00", "01-02", "03-04", "05-06", "07-08", "09-10", "11-12", "13-14", "15-16")
path_to_lPs <- "testingAllPhases"
map <- read.xls("NHANEScodes_ZS_2020-12-15.xlsx", as.is = TRUE, sheet = 3)
```

## Obtain desired data from saved output of bayesmarker runs  
Select the most recent cohort for each parent chemical from the full set of individual cohort runs.
If this data is not already available, users can alter their input file to include just the latest
cohort row for each individual biomarker before working through the bayesmarker pipeline (see the 
vignette calculateExposureInferences for the pipeline).

```{r readLPs}
# Subset the types of population groups here if desired
SUBPOP <- "all"
if (SUBPOP != "all"){
  ind <- SUBPOP %in% subpops
  subpops <- subpops[ind]
}

# Read in the exposure estimates for each cohort
dta.2 <- c()
for (j in 1:length(phases)){

  # Obtain file names for lPsamps
  if (is.null(path_to_lPs)){
    files <- list.files(path = ".", pattern = "lPsamps-gm_kg_day")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  } else {
    path <- file.path(".", path_to_lPs, phases[j])
    files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Read in files and build data.frame
  for (i in 1:length(subpops)){
    if (length(grep(paste("_", subpops[i], sep = ""), files)) == 0){
      # For cases like "0 - 5" supbop not in other phases.
      # Ok to call dta.1 here b/c 0-5 is not 1st in subpops
      chems <- unique(dta.1$CAS)
      dta <- data.frame(matrix(data = NA, nrow = length(chems), ncol = 2))
      colnames(dta) <- c("CAS", subpops[i])
      row.names(dta) <- chems
      dta.1 <- merge(dta.1, dta, by = "CAS", all.x = TRUE)
    } else {
      load(files[grep(paste("_", subpops[i], sep = ""), files)])
      samps <- as.matrix(lPsampsgm)
      result.med <- apply(samps, 2, mean)
      result.bar <- mean(apply(samps, 1, mean))
      dta <- data.frame(colnames(samps), exp(result.med))
      colnames(dta) <- c("CAS", subpops[i])
      if (i > 1){
        dta.1 <- merge(dta.1, dta, by = "CAS", all.x = TRUE)
      } else {
        dta.1 <- dta
      }
    }
  }
  dta.2 <- rbind(cbind(dta.1, "Phase" = rep(phases[j], dim(dta.1)[1])), dta.2)
}

# Get the most recent estimate for each chemical
parents <- as.character(unique(dta.2$CAS))
dta.2$CAS <- as.character(dta.2$CAS)
dta.2$Phase <- as.character(dta.2$Phase)
# Function to get the newest row for each chemical
getNewestPred <- function(parent, phases, data){
  availphase <- as.character(unique(dta.2$Phase[dta.2$CAS == parent]))
  ind <- max(match(availphase, phases))
  output <- dta.2[dta.2[,"CAS"] == parent & dta.2[,"Phase"] == phases[ind],]
  return(output)
}
final <- as.data.frame(t(sapply(parents, getNewestPred, phases = phases, data = dta.2)))
# Add preferred name and DTXSID
indP <- match(final$CAS, map$CAS)
final <- cbind("DTXSID" = map$DTXSID[indP], "Name" = map$Name[indP], final)
final <- apply(final, 2, as.character)
write.table(final, file = paste("ParentExpEsts_MostRecent_", format(Sys.time(), "%Y-%m-%d"), ".txt", sep = ""),
            quote = FALSE, sep = "\t", row.names = FALSE)
```

## Prepare main data object
```{r dataPrep}
#lPs <- read.delim(file = "ParentExpEsts_MostRecent_2021-06-23.txt", header = TRUE, sep = "\t")
lPs <- final
colnames(lPs)[-c(1:3,15)] <- c("Total", "Males", "Females", "Age.0.5", "Age.6.11", "Age.12.19", "Age.20.65", 
                               "Age.GT65", "ReproAgeFemale", "BMI <= 30", "BMI > 30")

# Add chemical class
map <- map[map$CAS %in% lPs$CAS,]  # 260 rows
map <- map[,c(2,3,4,7,8,9)]
classTable <- read.xls("allPhasesAllChems_ChemClass.xlsx", sheet = 1)
map$CAS.1 <- gsub(" ", "", map$CAS.1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])

# Shorten the class names for a concise plot
map$ChemClass <- as.character(map$ChemClass)
map$ChemClass[map$ChemClass == "Carbamate Pesticide Metabolites"] <- "Carbamate Pesticides"
map$ChemClass[map$ChemClass == "Flame Retardant Metabolite"] <- "Flame Retardant"
map$ChemClass[map$ChemClass == "Fungicides and Metabolites"] <- "Fungicides"
map$ChemClass[map$ChemClass == "Herbicides and Metabolites"] <- "Herbicides"
map$ChemClass[map$ChemClass == "Insect Repellent and Metabolites"] <- "Insect Repellent"
map$ChemClass[map$ChemClass == "Organochlorine Pesticide Metabolites"] <- "Organochlorine Pesticides"
map$ChemClass[map$ChemClass == "Personal Care and Consumer Product Chemicals"] <- "Personal Care/Consumer Product"
map$ChemClass[map$ChemClass == "Phthalate and Phthalate Alternative Metabolites"] <- "Phthalates"
map$ChemClass[map$ChemClass == "Phytoestrogens and Metabolites"] <- "Phytoestrogens"
map$ChemClass[map$ChemClass == "Polycyclic Aromatic Hydrocarbon Metabolites"] <- "Polycyclic Aromatic Hydrocarbons"
map$ChemClass[map$ChemClass == "Pyrethroid Metabolites"] <- "Pyrethroids"
map$ChemClass[map$ChemClass == "Volatile Organic Compound Metabolites"] <- "Volatile Organic Compounds"

tmp <- as.character(map$ChemClass[match(lPs$CAS, map$CAS)])
na.ind <- is.na(tmp)  # Only 3 
tmp <- tmp[!na.ind]
lPs <- lPs[!na.ind,]

lPs$Name <- substr(lPs$Name, 0, 25)

df <- as.data.frame(matrix(data = NA, nrow = dim(lPs)[1], ncol = 11, dimnames = list(lPs$Name, colnames(lPs)[4:14])))
# Calculate difference from total
for (i in 4:14){
  df[,i-3] <- log2(lPs[,i]) - log2(lPs$Total)
}

# Remove chemicals with very high uncertainty (due to high complexity metabolism or metabolites with most
# measurements below the LOD)
bigdiff <- unlist(apply(df, 1, function(x) (sum(x > 0, na.rm = TRUE) > 9)))  # 4 chemicals same sign for all pops
# Add to this those chemicals that are clearly troublesome
badChems <- c(77, 138, 23, 87, 88, 2, 169, 149, 165)
bigdiff[badChems] <- TRUE
df <- df[!bigdiff,]
tmp <- tmp[!bigdiff]
lPs <- lPs[!bigdiff,]

# Remove total population group b/c it adds no info to the plot
df$Total <- NULL
lPs$Total <- NULL

# Get cluster order for rows and columns
cluster <- "Class"  # Group by chemical class rather than numerical clustering
df2 <- df
# Replace missing data with median across all chemicals to allow for clustering
for (i in 1:dim(df2)[2]){
  df2[is.na(df2[,i]),i] <- median(as.numeric(df2[,i]), na.rm = TRUE)
}
# For rows 
if (cluster == "Class"){
  df <- df[order(tmp),]
} else if (cluster == "Ests"){
  dists <- dist(df2)
  dists <- hclust(dists, method = "complete") # contains the ordering of the rows
  df <- df[dists$order,]
}
# For columns
dists <- dist(t(df2))
dists <- hclust(dists, method = "complete") # contains the ordering of the columns
df <- df[,dists$order]
```

## Prepare plot aspects
```{r prepPlot}
row.anno <- data.frame("ChemicalClass" = tmp, row.names = lPs$Name)
cols <- primary.colors(length(unique(row.anno$ChemicalClass)))
anno.cols <- cols[1:length(unique(row.anno$ChemicalClass))]
names(anno.cols) <- levels(row.anno$ChemicalClass)
anno.cols <- list(ChemicalClass = anno.cols)

# To get value of 0 to be white (i.e. scale color palette)
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(min(df, na.rm = TRUE), 0, length.out = ceiling(paletteLength/2) + 1),
              seq(max(df, na.rm = TRUE)/paletteLength, max(df, na.rm = TRUE), length.out = floor(paletteLength/2)))

# Set up gaps by class: anno.cols is in the correct order, so step through this and keep a running
# count of how many is in each class to set up a gaps vector
gaps <- numeric(length(anno.cols$ChemicalClass) - 1)
k <- 0
for (i in 1:(length(anno.cols$ChemicalClass)-1)){
  temp <- sum(row.anno$ChemicalClass == names(anno.cols$ChemicalClass)[i])
  gaps[i] <- k + temp
  k <- k + temp 
}
```

## Make the plot
```{r plot}
p <- pheatmap(df, color = myColor, breaks = myBreaks, border_color = "gray",
          cluster_cols = TRUE, cluster_rows = FALSE, annotation_row = row.anno,
          main = "Difference in Estimated Exposure Across Population Groups", fontsize = 12,
          fontsize_row = 12, annotation_colors = anno.cols, annotation_names_col = FALSE,
          show_rownames = TRUE, annotation_names_row = FALSE, na_col = "grey",
          annotation_legend = TRUE, show_colnames = TRUE, fontsize_col = 15, gaps_row = gaps)
p <- p + grid.text("Log2(Group-Total)", x = 0.68, y = 0.964)

pdf(file = paste("log2ExpEstvsTotalHeatmap_all_vert_red_legend_", format(Sys.time(), "%Y-%m-%d"), ".pdf", sep = ""),
    width = 12.5, height = 25)
  print(p)
dev.off()

```


## Save supplemental data file for this plot
```{r saveData}
ind <- match(row.names(df), row.names(row.anno))
df$ChemicalClass <- row.anno$ChemicalClass[ind]
df$Name <- row.names(df)
ind <- match(df$Name, lPs$Name)
df$CAS <- lPs$CAS[ind]
df$DTXSID <- lPs$DTXSID[ind]
df <- df[,c(14,13,12,11,5,1,2,9,10,7,3,8,4,6)]
df[,5:14] <- round(df[,5:14], digits = 4)
write.table(df, file = "bayesmarkerPaper_Fig4_data.txt", sep = "\t", row.names = FALSE)
```


















