---
title: "Compare Exposure Trends Across All Population Groups for a Set of Chemicals"
author: "Zachary Stanfield"
date: "2/23/2022"
output: html_document
---

from "Characterizing Exposure Trends from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This document recreates figures 4, 5, and 6 from the paper.  It uses the Markov chains calculated
from the main bayesmarker pipeline (see the vignette calculateExposureInferences.Rmd) to plot 
individual trend lines (with median exposure and 95% CI for each point) for all population groups 
for individual chemicals.

## Prepare for session 
### Load relevant libraries
```{r setup}
library(readxl)
library(reshape2)
library(stringr)
library(ggplot2)
```

### Declare some needed objects
```{r objs}
subpops <- c("Total", "Male", "Female", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
             "ReproAgeFemale", "BMI <= 30", "BMI > 30")
names(subpops) <- c("Total", "Male", "Female", "6 - 11 years", "12 - 19 years", "20 - 65 years", "66 years and older",
                    "ReproAgeFemale", "BMI_le_30", "BMI_gt_30")
phases <- c("99-00", "01-02", "03-04", "05-06", "07-08", "09-10", "11-12", "13-14", "15-16")
path_to_lPs <- "/ccte/home1/zstanfie/MetaboliteInference/testingAllPhases"
phthalates <- c("84-66-2", "117-81-7", "28553-12-0", "84-69-5")
names(phthalates) <- c("Diethyl phthalate", "Di-2-ethylhexyl phthalate", "Di-isononyl phthalate", "Di-isobutyl phthalate")
parabens <- c("120-47-8", "94-13-3", "94-26-8", "99-76-3")
names(parabens) <- c("Ethyl paraben", "n-propyl paraben", "Butyl paraben", "Methyl paraben")
others <- c("7440-43-9", "64436-13-1", "68-12-2", "107-13-1")
names(others) <- c("Cadmium", "Arsenobetaine", "N,N-Dimethylformamide", "Acrylonitrile")
```

## Retrieve Markov chains and organize into dataframs
```{r getlPs}
# Use 1 data object for each final plot point item (median exposure, upper value 
# of the 95% CI, and lower value)
dta.2 <- c()
dta.2.low <- c()
dta.2.up <- c()
for (j in 1:length(phases)){
  
  # Obtain file names for lPsamps
  if (is.null(path_to_lPs)){
      files <- list.files(path = ".", pattern = "lPsamps-gm_kg_day")
  } else {
    path <- file.path(path_to_lPs, phases[j])
      files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
      dates <- "2021-05-11"
      files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Read in files and build data.frame
  for (i in 1:length(subpops)){
    if (length(grep(paste("_", subpops[i], sep = ""), files)) == 0){
      # For cases when data is missing for the supbop 
      # Ok to call dta.1 here b/c data is never missing for the first subpop
      chems <- unique(dta.1$CAS)
      dta <- data.frame(matrix(data = NA, nrow = length(chems), ncol = 2))
      colnames(dta) <- c("CAS", subpops[i])
      row.names(dta) <- chems
      dta.1 <- merge(dta.1, dta, by = "CAS", all.x = TRUE)
    } else {
      load(files[grep(paste("_", subpops[i], sep = ""), files)])
      samps <- as.matrix(lPsampsgm)
      # Having all NAs for a column causes an error
      rmind <- unlist(apply(samps, 2, function(x) sum(is.na(x)) == length(x)))
      samps <- samps[,!rmind]
      result.med <- apply(samps, 2, mean)
      result.bar <- mean(apply(samps, 1, mean))
      # Get 95th Percentile
      result.CI <- t(apply(samps, 2, function(x) quantile(x, probs = c(0.0275, 0.975))))
      dta <- data.frame(colnames(samps), exp(result.med))
      dta.low <- data.frame(colnames(samps), exp(result.CI[,1]))
      dta.up <- data.frame(colnames(samps), exp(result.CI[,2]))
      colnames(dta) <- c("CAS", subpops[i])
      colnames(dta.low) <- c("CAS", paste(subpops[i], "_low95", sep = ""))
      colnames(dta.up) <- c("CAS", paste(subpops[i], "_up95", sep = ""))
      if (i > 1){
        dta.1 <- merge(dta.1, dta, by = "CAS", all.x = TRUE)
        dta.1.low <- merge(dta.1.low, dta.low, by = "CAS", all.x = TRUE)
        dta.1.up <- merge(dta.1.up, dta.up, by = "CAS", all.x = TRUE)
      } else {
        dta.1 <- dta
        dta.1.low <- dta.low
        dta.1.up <- dta.up
      }
    }
  }
  dta.2 <- rbind(cbind(dta.1, "Phase" = rep(phases[j], dim(dta.1)[1])), dta.2)
  dta.2.low <- rbind(cbind(dta.1.low, "Phase" = rep(phases[j], dim(dta.1.low)[1])), dta.2.low)
  dta.2.up <- rbind(cbind(dta.1.up, "Phase" = rep(phases[j], dim(dta.1.up)[1])), dta.2.up)
}
```

## Reduce data to just the chemicals we will eventually plot
```{r redChems}
dta.2 <- dta.2[dta.2$CAS %in% c(phthalates, parabens, others),]
dta.2.low <- dta.2.low[dta.2.low$CAS %in% c(phthalates, parabens, others),]
dta.2.up <- dta.2.up[dta.2.up$CAS %in% c(phthalates, parabens, others),]
```


## Coerce data into a set of CASxCohort matrices for each population group
This will make them easier to work with later on.  Impute missing values to help visualization
of exposure trend changes.  Keep track of which values are imputed so we can indicate them
in the final plot.
```{r buildDFs}
alldfs <- list()
impdfs <- list()
for (k in 1:length(subpops)){
  # Extract subpop matrix
  df <- dta.2[,colnames(dta.2) %in% c("CAS", "Phase", unname(subpops[k]))]
  df <- tapply(df[,2], list(df$CAS, df$Phase), function(x) mean(x, na.rm = TRUE))
  df <- df[,c(9,1:8)]
  indNaN <- is.nan(df)
  df[indNaN] <- NA
  df <- log(df)
  
  # Keep track of imputed values
  impdfs[[k]] <- df
  impdfs[[k]][is.na(df)] <- "Yes"
  impdfs[[k]][!(df == "Yes")] <- "No"
  impdfs[[k]] <- melt(impdfs[[k]])
  colnames(impdfs[[k]]) <- c("Chemical", "Cohort", "Imputed")
  
  # Value imputation
  for (i in 1:dim(df)[1]){  # Step through each chemical
    allVals <- which(!is.na(df[i,]))
    firstVal <- allVals[1]
    lastVal <- allVals[length(allVals)]
    if (length(allVals) == 0){    # When all data is missing (either not measured in cohorts or estimate is NA)
      df[i,] <- NA
      next
    }
    if (firstVal != 1){ # Use the value for the first earliest non-missing cohort for all those preceeding
      df[i, 1:(firstVal - 1)] <- df[i,firstVal]  # in other words, propogate first measured value backwards
    }
    if (lastVal != dim(df)[2]){ # propogate the last value forwards
      df[i, (lastVal + 1):dim(df)[2]] <- df[i,lastVal]
    }
    miss <- setdiff(seq(firstVal, lastVal), allVals)
    
    for (j in miss){
      below <- max(which((j - allVals) > 0))  # index with closest previous value
      above <- allVals[Position(function(x) x > j, allVals)]  # index with closest newer value
      df[i, j] <- mean(df[i,c(below,above)])  # use the mean of these two as the imputed value
    }
  }
  
  # Record the df
  alldfs[[k]] <- df 
  
  # Ensure df and dftot have matching chemicals
  #tmpDF <- dftot[row.names(dftot) %in% row.names(df),]
  
}
names(alldfs) <- subpops
```


# Need to create a list of dfs for both upper and lower CI, like alldfs above
```{r CIdfs}
alldfs.low <- list()
for (i in 1:length(subpops)){
  df <- dta.2.low[,colnames(dta.2.low) %in% c("CAS", "Phase", paste(unname(subpops[i]), "_low95", sep = ""))]
  df <- tapply(df[,2], list(df$CAS, df$Phase), function(x) mean(x, na.rm = TRUE))
  df <- df[,c(9,1:8)]
  indNaN <- is.nan(df)
  df[indNaN] <- NA
  df <- log(df)
  alldfs.low[[i]] <- df
}
names(alldfs.low) <- names(alldfs)

alldfs.up <- list()
for (i in 1:length(subpops)){
  df <- dta.2.up[,colnames(dta.2.up) %in% c("CAS", "Phase", paste(unname(subpops[i]), "_up95", sep = ""))]
  df <- tapply(df[,2], list(df$CAS, df$Phase), function(x) mean(x, na.rm = TRUE))
  df <- df[,c(9,1:8)]
  indNaN <- is.nan(df)
  df[indNaN] <- NA
  df <- log(df)
  alldfs.up[[i]] <- df
}
names(alldfs.up) <- names(alldfs)
```


## Create main plot data frame by combining all the subpop data frames
```{r getPlotDF}
# Switch to using chemical names rather than CAS
map <- read_excel("/ccte/home1/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 3)
tomatch <- row.names(alldfs$Total)
indName <- match(tomatch, map$CAS)
indNames <- map$Name[indName]

# Loop by subpop to add all data to the same object
pdf_final <- c()
for (i in 1:length(subpops)){
  pdf1 <- data.frame("Chemical" = indNames, "temp" = alldfs[[i]])
  pdf1.low <- data.frame("Chemical" = indNames, "lower95" = alldfs.low[[i]])
  pdf1.up <- data.frame("Chemical" = indNames, "upper95" = alldfs.up[[i]])
  colnames(pdf1) <- gsub("temp", names(subpops)[i], colnames(pdf1))
  pdf1 <- melt(pdf1)
  pdf1.low <- melt(pdf1.low)
  pdf1.up <- melt(pdf1.up)
  sepcols <- str_split_fixed(pdf1$variable, "\\.", 2)
  pdf1 <- cbind(pdf1[,c(1,3)], sepcols, pdf1.low$value, pdf1.up$value)
  colnames(pdf1) <- c("Chemical", "Exposure", "Population", "Cohort", "lower95", "upper95")
  pdf1$Exposure <- exp(pdf1$Exposure)
  pdf1$lower95 <- exp(pdf1$lower95)
  pdf1$upper95 <- exp(pdf1$upper95)
  
  # Need to concatenate the imputed values as well to add to the final df
  imptemp_col <- c()
  for (j in 1:length(phases)){
    imptemp <- impdfs[[i]][impdfs[[i]]$Cohort == phases[j],]
    imptemp_col <- c(imptemp_col, imptemp$Imputed)
  }
  pdf1$Imputed <- imptemp_col
  
  # We can now contribute the df for this population to the final df
  pdf_final <- rbind(pdf_final, pdf1)
}
```


## Setting the df factors up
```{r factors}
pdf_final$Chemical <- factor(pdf_final$Chemical, levels = map$Name[indName])
pdf_final$Cohort <- as.factor(pdf_final$Cohort)
pdf_final$Cohort <- factor(pdf_final$Cohort, levels = levels(pdf_final$Cohort)[c(9,1:8)])
pdf_final$Imputed <- as.factor(pdf_final$Imputed)
levels(pdf_final$Imputed) <- c("No", "Yes")
pdf_final$Population <- as.factor(pdf_final$Population)
pdf_final$Population <- factor(pdf_final$Population, levels = levels(pdf_final$Population)[c(10,8,7,3,1,2,4,9,6,5)])
```


## Set some plot elements
```{r plotParams}
myColors <- c("Black", rainbow(length(levels(pdf_final$Population)) - 1))
names(myColors) <- levels(pdf_final$Population)
alphas <- c("Total" = 1, "Male" = 0.5, "Female" = 0.5, "6 - 11 years" = 0.5, "12 - 19 years" = 0.5, 
            "20 - 65 years" = 0.5, "66 years and older" = 0.5,
            "ReproAgeFemale" = 0.5, "BMI <= 30" = 0.5, "BMI > 30" = 0.5)
```


## Make the plots
```{r makePlots, echo=TRUE}
# Phthalates
pht_df <- pdf_final[pdf_final$Chemical %in% names(phthalates),]
pht_df$Chemical <- factor(pht_df$Chemical, levels = names(phthalates))
p <- ggplot(pht_df, aes(x = Cohort, y = Exposure, group = Population, color = Population, alpha = Population)) +
  geom_point(aes(shape = Imputed), size = 2.25) +
  scale_shape_manual(values = c(19,1)) +
  geom_line() +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, position=position_dodge(0.25)) +
  scale_color_manual(name = "Population", values = myColors) +
  scale_alpha_manual(values = alphas) +
  scale_y_log10() +
  ylab("Exposure (mg/kg/day)") +
  ggtitle("Phthalate Exposure Trends") +
  facet_wrap(.~Chemical, nrow = 2, scales = "free_y") +
  theme_set(theme_gray(base_size = 14)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  guides(size = "legend", alpha = "none")
print(p)
pdf(file = "SubpopExposureTrend_Phthalates.pdf", width = 13, height = 11)
  print(p)
dev.off()

# Parabens
bens_df <- pdf_final[pdf_final$Chemical %in% names(parabens),]
bens_df$Chemical <- factor(bens_df$Chemical, levels = names(parabens))
p <- ggplot(bens_df, aes(x = Cohort, y = Exposure, group = Population, color = Population, alpha = Population)) +
  geom_point(aes(shape = Imputed), size = 2.25) +
  scale_shape_manual(values = c(19,1)) +
  geom_line() +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, position=position_dodge(0.25)) +
  scale_color_manual(name = "Population", values = myColors) +
  scale_alpha_manual(values = alphas) +
  scale_y_log10() +
  ylab("Exposure (mg/kg/day)") +
  ggtitle("Parabens Exposure Trends") +
  facet_wrap(.~Chemical, nrow = 2, scales = "free_y") +
  theme_set(theme_gray(base_size = 14)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  guides(size = "legend", alpha = "none")
print(p)
pdf(file = "SubpopExposureTrend_Parabens.pdf", width = 13, height = 11)
  print(p)
dev.off()

# Others
others_df <- pdf_final[pdf_final$Chemical %in% names(others),]
others_df$Chemical <- factor(others_df$Chemical, levels = names(others))
p <- ggplot(others_df, aes(x = Cohort, y = Exposure, group = Population, color = Population, alpha = Population)) +
  geom_point(aes(shape = Imputed), size = 2.25) +
  scale_shape_manual(values = c(19,1)) +
  geom_line() +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, position=position_dodge(0.25)) +
  scale_color_manual(name = "Population", values = myColors) +
  scale_alpha_manual(values = alphas) +
  scale_y_log10() +
  ylab("Exposure (mg/kg/day)") +
  ggtitle("Atypical Exposure Trends") +
  facet_wrap(.~Chemical, nrow = 2, scales = "free_y") +
  theme_set(theme_gray(base_size = 14)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  guides(size = "legend", alpha = "none")
print(p)
pdf(file = "SubpopExposureTrend_Workplace.pdf", width = 13, height = 11)
  print(p)
dev.off()

```




