---
title: "Make Heatmaps Showing Exposures Over NHANES Cohorts"
author:  "Zachary Stanfield"
date:  "February 11, 2022"
output: html_document
---

from "Characterizing Exposure Trends from NHANES Urine Biomonitoring Data"

Zachary Stanfield, R. Woodrow Setzer, Victoria Hull, Risa R. Sayre, Kristin K. Isaacs, John F. Wambaugh

(Insert Journal Name, issue, data, and page numbers)

(Insert link to DOI)

# Description
This file uses the inferred exposures for parent chemicals of NHANES biomarkers that was obtained using 
the bayesmarker package to generate a heatmap of median exposure across all survey participants. A 
heatmap is also generated for the metabolite "exposures".

## Prepare for session 
### Load relevant libraries
```{r setup}
library(readxl)
library(RColorBrewer)
library(colorRamps)
library(pheatmap)
```

## Generate Parent Chemical Heatmap
### Declare some needed objects
```{r dataObjs}
subpops <- "Total"
phases <- c("99-00", "01-02", "03-04", "05-06", "07-08", "09-10", "11-12", "13-14", "15-16")
path_to_lPs <- "/ccte/home1/zstanfie/MetaboliteInference/testingAllPhases"
map <- as.data.frame(read_excel("/ccte/home1/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 3))
```


### Obtain desired data from saved output of bayesmarker runs for generating the parent chemical heatmap
Retrieve all cohort Markov chains for each parent chemical for the total population.
See the vignette calculateExposureInferences.Rmd to generate the Markov chains if desired.
```{r readlPs}
dta.2 <- c()
for (j in 1:length(phases)){

  # Obtain file names for lPsamps
  if (is.null(path_to_lPs)){
    files <- list.files(path = ".", pattern = "lPsamps-gm_kg_day")
    # If you have results from multiple runs, you can select from a specific day like this
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  } else {
    path <- file.path(path_to_lPs, phases[j])
    files <- paste(path, "/", list.files(path = path, pattern = "lPsamps-gm_kg_day"), sep = "")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Read in files and build data.frame
  load(files[grep(paste("_", subpops, sep = ""), files)])
  samps <- as.matrix(lPsampsgm)
  result.med <- apply(samps, 2, mean)
  result.bar <- mean(apply(samps, 1, mean))
  dta <- data.frame(colnames(samps), exp(result.med))
  colnames(dta) <- c("CAS", subpops)
  
  dta.2 <- rbind(cbind(dta, "Phase" = rep(phases[j], dim(dta)[1])), dta.2)
}

# Get in the right format
dfT <- as.data.frame(with(dta.2[,colnames(dta.2) %in% c("CAS", "Phase", "Total")], 
            tapply(Total, list(CAS, Phase), function(x) mean(x, na.rm = TRUE))))
```
   
            
### Add chemical class
```{r nameParents}
map <- map[map$CAS %in% row.names(dfT),]  
map <- map[,c(2,3,4,7,8,9)]
classTable <- as.data.frame(read_excel("/ccte/home1/zstanfie/MetaboliteInference/allPhasesAllChems_ChemClass.xlsx", sheet = 1))
map$CAS.1 <- gsub(" ", "", map$CAS.1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])
```


### Cluster exposure profiles and save plot 
```{r parentPlot}
df <- dfT[,c(9,1:8)]  
df <- log(df)
df2 <- df
# There missing values in our data frame, but this is not allowed when clustering
# Impute missing values with the median exposure over all chemicals for a cohort
for (i in 1:dim(df2)[2]){
  df2[is.na(df2[,i]), i] <- median(df2[,i], na.rm = TRUE)
}
dists <- dist(df2, method = "euclidean")
dists <- hclust(dists, method = "complete") # contains the ordering of the rows
df <- df[dists$order,]

# Create row annotation for the chemical class
tmp <- as.character(map$ChemClass[match(row.names(df), map$CAS)])
row.anno <- data.frame("ChemicalClass" = tmp, row.names = row.names(df))
row.anno$ChemicalClass <- as.factor(row.anno$ChemicalClass) 
cols <- primary.colors(length(unique(row.anno$ChemicalClass)))
anno.cols <- cols[1:length(unique(row.anno$ChemicalClass))]
names(anno.cols) <- levels(row.anno$ChemicalClass)
anno.cols <- list(ChemicalClass = anno.cols)
pheatmap(df, color = colorRampPalette(brewer.pal(n = 7, name = "YlOrRd"))(100),
         cluster_cols = FALSE, cluster_rows = FALSE, annotation_row = row.anno,
         main = "Parent Chemicals Log of Median Exposure Estimates (mg/kg/day) by Cohort", fontsize = 18,
         fontsize_row = 8, fontsize_col = 19, angle_col = 0, annotation_colors = anno.cols,
         annotation_names_row = FALSE, na_col = "grey", 
         filename = paste("ExpEstHeatmapEuclid_all_", format(Sys.time(), "%Y-%m-%d"), ".pdf", sep = ""),
         width = 19, height = 17, silent = TRUE)
```         
         
         
         
## Generate Metabolite Heatmap
### Declare data objects and obtain the Markov chains for the metabolites
```{r readlUs}
path_to_lUs <- "/ccte/home1/zstanfie/MetaboliteInference/testingAllPhases"
map <- read_excel("/ccte/home1/zstanfie/MetaboliteInference/NHANEScodes_ZS_2020-12-15.xlsx", sheet = 3)

dta.m <- c()
for (j in 1:length(phases)){
  
  # Obtain file names for lUs
  if (is.null(path_to_lUs)){
    files <- list.files(path = ".", pattern = "lU-gm_kg_day")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  } else {
    path <- file.path(path_to_lUs, phases[j])
    files <- paste(path, "/", list.files(path = path, pattern = "lU-gm_kg_day"), sep = "")
    dates <- "2021-05-11"
    files <- files[unlist(sapply(dates, function(x) grep(x, files)))]
  }
  
  # Read in files and build data.frame
  load(files[grep(paste("_", subpops, sep = ""), files)])
  samps <- as.matrix(lU)
  result.med <- apply(samps, 2, mean)
  result.bar <- mean(apply(samps, 1, mean))
  dta <- data.frame(colnames(samps), exp(result.med))
  colnames(dta) <- c("CAS", subpops)
      
  dta.m <- rbind(cbind(dta, "Phase" = rep(phases[j], dim(dta)[1])), dta.m)
}

dfTm <- as.data.frame(with(dta.m[,colnames(dta.m) %in% c("CAS", "Phase", "Total")], 
            tapply(Total, list(CAS, Phase), function(x) mean(x, na.rm = TRUE))))
```


### Add class lables
```{r nameMetabs}
map <- map[map$CAS.1 %in% row.names(dfTm),]  
map <- map[,c(2,3,4,7,8,9)]
classTable <- read_excel("/ccte/home1/zstanfie/MetaboliteInference/allPhasesAllChems_ChemClass.xlsx", sheet = 1)
map$CAS.1 <- gsub(" ", "", map$CAS.1)
map$ChemClass <- apply(map, 1, function(x) classTable$Column4[match(x[5], classTable$Column1)])
```


### Cluster exposure profiles and save plot
```{r metabPlot}
df <- dfTm[,c(9,1:8)]
df <- log(df)
df2 <- df
# There missing values in our data frame, but this is not allowed when clustering
# Impute missing values with the median exposure over all chemicals for a cohort
for (i in 1:dim(df2)[2]){
  df2[is.na(df2[,i]), i] <- median(df2[,i], na.rm = TRUE)
}
dists <- dist(df2)
dists <- hclust(dists, method = "complete") # contains the ordering of the rows
df <- df[dists$order,]

# Create row annotation for the chemical class
tmp <- as.character(map$ChemClass[match(row.names(df), map$CAS.1)])
row.anno <- data.frame("ChemicalClass" = tmp, row.names = row.names(df))
row.anno$ChemicalClass <- as.factor(row.anno$ChemicalClass) 
cols <- primary.colors(length(unique(row.anno$ChemicalClass)))
anno.cols <- cols[1:length(unique(row.anno$ChemicalClass))]
names(anno.cols) <- levels(row.anno$ChemicalClass)
anno.cols <- list(ChemicalClass = anno.cols)
pheatmap(df, color = colorRampPalette(brewer.pal(n = 7, name = "YlOrRd"))(100),
         cluster_cols = FALSE, cluster_rows = FALSE, annotation_row = row.anno,
         main = "Metabolite Log Mean Concentration (mg/kg/day) by Cohort", fontsize = 17,
         fontsize_row = 8, angle_col = 0, annotation_colors = anno.cols,
         annotation_names_row = FALSE, na_col = "grey", 
         filename = paste("ExpEstHeatmapEuclid_allMetabs_", format(Sys.time(), "%Y-%m-%d"), ".pdf", sep = ""),
         width = 16, height = 16, silent = TRUE)
```

